---
title: "Word frequency analysis"
teaching: 0
exercises: 0
---


:::::::::::::::::::::::::::::::::::::: questions 

- How is a frequency analysis conducted?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn how to find frequent words
- Learn how to analyse and visualise it


::::::::::::::::::::::::::::::::::::::::::::::::



```{r load_libraries, include = FALSE}
library(tidyverse)
library(tidytext)
library(tm)

articles_anti_join <- read_csv("data/articles_anti_join.csv", na = c("NA", "NULL", ""))
articles <- read_csv("data/guardianArticles.csv", na = c("NA", "NULL", ""))
```


## Frequency analysis

A word frequency is a relatively simple analysis. It measures how often words occur in a text. 


```{r simple_frequency_count}
articles_anti_join |> 
  count(word, sort = TRUE)
```

The previous code chunk resulted in a list containing the most frequent words. The words are from articles about both presidents, and they are sorted based on frequency with the highest number on top.

A closer look at the list may reveal that some words are irrelevant. Given that the articles in the dataset are about the two presidents' respective inaugurations, we consider the words below irrelevant for our analysis. Therefore, we make a new dataset without these words.

```{r filtered_frequency_count}
articles_filtered <- articles_anti_join |> 
  filter(!word %in% c("it’s", "don’t"))

articles_filtered |> 
  count(word, sort = TRUE)
```
The words deemed irrelevant are no longer on the list above.

Instead of a general list it may be more interesting to focus on the most frequent words belonging to articles about the two presidents respectively.

```{r count_pr_president}
articles_filtered |> 
  count(section, word, sort = TRUE)
```

Keeping an overview of the words associated with each section can be a bit tricky. For instance, the word "time" is associated with both lifestyle and sport. This is easy to see, as the two words are right next to each other, but what if the words are further apart.


<!-- ```{r top_ten_words_pr_president} -->
 <!-- articles_filtered |> 
   count(section, word, sort = TRUE) |> 
   group_by(section) |> 
   slice(1:5) |> 
   ggplot(mapping = aes(x = n, y = word, colour = section, shape = section)) +
   geom_point()  -->
<!-- ``` -->

<!-- The plot above shows the top-five words associated with the sections respectively. If a word features on multiple sections' top-five list, it only occurs once in the plot. This is why the plot doesn't contain 35 words in total. -->

Another way of looking at the words compared with eachother in section could be a table where we count the words used pr sections easily comparable. In this analysis the section is the guiding principle. 



```{r}
articles_filtered |> 
  count(section, word, sort = TRUE) |> 
  pivot_wider(
    names_from = section,
    values_from = n
  )
```

# Proportion
It is well enough to count the words, but we do not know wether 3131 word about AI in the news section is alot compared to 1250 word in the opnion section if we compare to the total amount of words used in each section.

So let us have a look at how it looks if we make the same query, but with proportion insted of word count.

```{r proportions}
articles_filtered |> 
  count(section, word) |> 
  group_by(section) |> 
  mutate(proportion = n / sum(n)) |> 
  ungroup() |> 
  select(-n) |> 
  arrange(desc(proportion)) |>
  pivot_wider(
    names_from = section,
    values_from = proportion)
```


# Visualization

We can also visualize word frequency. In the following we will visualize what the top 10 word used in each section is

```{r top_ten_visusalised}
articles_filtered |> 
  group_by(section) |> 
  count(word, sort = TRUE) |> 
  top_n(10) |> 
  ungroup() |> 
  mutate(word = reorder_within(word, n, section)) |> 
  ggplot(aes(n, word, fill = section)) +
  geom_col() +
  facet_wrap(~section, scales = "free") +
  scale_y_reordered() + 
  labs(x = "word occurrences")

```

<!-- The analyses just made can easily be adjusted. For instance, if we want look at the words by `pillar_name` instead of by `president`, we simply replace `president` with `pillar_name` in the code. -->

<!-- ```{r vis_per_pillar} -->
<!-- articles_filtered |> 
  count(date, word, sort = TRUE) |> 
  group_by(date) |> 
  slice(1:10) |> 
  ggplot(mapping = aes(x = n, y = word, colour = date, shape = date)) +
  geom_point()  -->
<!-- ``` -->


Interesting that numbers occurs in the Sports section. Lets have a look at where it occurs. In order to do so we have to go back to our original object `articles` where we have the full text of articles. It would be nice to see the context in wich for instance 1 occurs.

In order to do this we need to use a `package` named `quanteda`, so lets install the package, and load it in to memory.

```{r, eval = FALSE}
install.packages("quanteda")
library(quanteda)
```

In order to get the context we will use the function `kwic`
```{r}
articles |> 
  filter(section == "Sport") |> 
  # filter(str_detect(text, "\\b1\\b")) |> 
  # slice(1) |> 
  pull(text) |> 
  tokens() |>
  kwic(pattern = "1", window = 10)

```

So what we are doing is looking for at keyword in context, and that is exactly what the function `kwic` does. It locates the string we write in the argument `pattern` and takes the amout of words before and after the keyword given in the argument `window`.



```{r saving_data, include = FALSE}
write_csv(articles_filtered, "data/articles_filtered.csv")
```


::::::::::::::::::::::::::::::::::::: keypoints 

- Making a frequency analysis
- Visualising the results


::::::::::::::::::::::::::::::::::::::::::::::::
