---
title: "Sentiment analysis"
teaching: 0
exercises: 0
---

:::::::::::::::::::::::::::::::::::::: questions 

- What is a sentiment?
- How is sentiment analysis conducted?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn about different lexicons
- Learn how to add sentiment to words using lexicons
- Analyse and visualise the sentiments in a text


::::::::::::::::::::::::::::::::::::::::::::::::

```{r chunk_options, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE)
```


```{r articles_filtered, include = FALSE}
library(tidyverse)
library(tidytext)
library(tm)
library(textdata)

articles_filtered <- read_csv("data/articles_filtered.csv", na = c("NA", "NULL", ""))
articles <- read_csv("data/guardianArticles.csv", na = c("NA", "NULL", ""))
articles_anti_join <- read_csv("data/articles_anti_join.csv", na = c("NA", "NULL", ""))
```

## Sentiment analysis

Sentiment refers to the emotion or tone in a text. It is typically categorised as positive, negative or neutral. Sentiment is often used to analyse opinions, attitudes or emotions in written content. In this case the written content is newspaper articles.

Sentiment analysis is a method used to identify and classify emotions in textual data. This is often done using word list (lexicons). The goal is to determine whether a given text has a positive, negative or neutral tone.

In order to do a sentiment analysis on our data we 
From the previous section we have a dataset containing a list of words in the text without stopwords. To do a sentiment analysis we can use a so-called lexicon and assign a sentiment to each word. In order to do this we need an list of words and their sentiment. A simple form would be wether they are positive or negative.

There are multiple sentiment lexicons. For a start we will be using the `bing` lexicon. This lexicon categorizes words as either positive or negative.


```{r get_sentiment_bing}
get_sentiments("bing")
```

In order to use the `bing`-lexicon, we have to save it.

```{r object_get_sentiment_bing}
bing <- get_sentiments("bing")
```

We now need to combine the sentiment to the words from our articles. We do this by performing an inner_join.

```{r bing_inner_join, warning = FALSE}
articles_bing <- articles_filtered |> 
  inner_join(bing)

articles_bing
```

In R, `inner_join()` is commonly used to combine datasets based on a shared column. In this case it is the `word` column. `inner_join()` matches words from a text dataset, in this case `articles_filtered` with words in the Bing sentiment lexicon to determine whether they are positive or negative.

When we have the combined dataset we can begin making a sentiment analysis. A start could be to count the number of positive and negative words used in articles, per president.

```{r articles_bing_group_by_inner_join}
articles_bing |> 
  group_by(section) |> 
  summarise(positive = sum(sentiment == "positive"),
            negative = sum(sentiment == "negative"),
            difference = positive - negative) 
```

This shows that more positive than negative words are associated with both presidents. It also shows that Trump is the president with the highest number of associated negative words.

```{r articles_bing_group_by_inner_join_graph}
articles_bing |> 
  group_by(section, date) |> 
  summarise(positive = sum(sentiment == "positive"),
            negative = sum(sentiment == "negative"),
            difference = positive - negative) |> 
  #ungroup() |> 
  filter(section %in% c("Arts", "Sport")) |> 
  ggplot(mapping = aes(x = date, y = difference, colour = section, group = section)) +
  geom_point() +
  geom_line()
 
```

Looking at the graphs we can see that the wording in december in sports articles are quite negative compared to february. If we had a data set covering more years it would be interesting to see if this was a normal thing. For now it might be interesting to read the articles from december and februar and compare what they are writing about. So how do we get the articles

```{r}
interesting_articles <- articles |> 
  filter(section == "Sport") |> 
  filter(date %in% c("2025-12", "2026-02"))
```

```{r eval = FALSE}
write_csv(interesting_articles, "data_out/interesting_articles.csv")
```


<!-- Another interesting thing to look at would the 10 most positive and negative words used in the articles. -->

<!-- ```{r facet_wrap} -->
<!-- # articles_bing |> 
  # count(word, sentiment, sort = TRUE) |> 
  # ungroup() |> 
  # group_by(sentiment) |> 
  # slice_max(n, n = 10) |> 
  # ungroup() |> 
  # mutate(word = reorder(word, n)) |> 
  # ggplot(mapping = aes(n, word, fill = sentiment)) +
  # geom_col(show.legend = FALSE) +
  # facet_wrap(~sentiment, scales = "free_y") -->

<!-- ``` -->

<!-- Here we can see the positive and negative words used in the articles. -->

With `bing` we only look at the sentiment in a binary fashion - a word is either positive or negative. If we try to do a similar analysis 
with `AFINN`, it looks different. `AFINN` is a sentiment lexicon. It consists of a list of words that are assigned sentiment scores ranging from -5 (very negative) to +5 (very positive).

`AFINN` is part of the package `textdata`, so we need to install the package and run library in order to be able to use it in this script.

```{r install_packages_textdata, eval = FALSE}
install.packages("textdata")
library(textdata)
```

In order to use the `AFINN`-lexicon, we have to save it.

```{r get_sentiment_afinn}
afinn <- get_sentiments("afinn")
```

Let's have a look at it.

```{r indlaes_afinn, echo = FALSE, message = FALSE}
afinn <- read_csv("data/AFINN.CSV")
```

```{r vis_afinn}
afinn
```


:::: instructor
Bemærk at vi ikke på github kan downloade afinn. Derfor 
har vi downloaded afinn datasættet til en csv-fil pr 21. november 2025.
Med andre ord er der risiko for at siden kører med et uopdateret
datasæt. - TROR IKKE VI BEHØVER DET MERE - SNAK MED CHRISTIAN
::::
We now need to combine the sentiment to the words from our articles. We do this by performing an inner_join.

```{r inner_join_afinn}
articles_afinn <- articles_filtered |> 
  inner_join(afinn) 
```

Since the `AFINN` lexicon adds negative and positive numbers to the words (instead of strings as Bing does) we can easily calculate the difference as we did with `bing`. In order to see wether a section is dominated by positive or negative words.

```{r afinn_summarise}
articles_afinn |> 
  group_by(section) |> 
  summarise(different = sum(value))
```

It could be interesting to see how the different levels of negative and positive words are used in the different sections. 

```{r}
articles_afinn |> 
  #group_by(section) |> 
  count(section, value) |> 
  pivot_wider(names_from = value, values_from = n)
```


```{r}
articles_afinn |> 
  count(section, value) |> 
  group_by(section) |>
  mutate(proportion = n / sum(n)) |> 
  ungroup() |> 
  select(-n) |> 
  arrange(desc(proportion)) |> 
  pivot_wider(names_from = value, values_from = proportion)
```

```{r}
articles_afinn |> 
  count(section, value) |> 
  group_by(section) |>
  mutate(proportion = n / sum(n)) |> 
  ungroup() |> 
  select(-n) |> 
  filter(section %in% c("Sport", "Opinion"))  |> 
  ggplot(mapping = aes(x = value, y = proportion, fill = section)) +
  geom_col(position = "dodge")
```

<!-- ```{r afinn_president_value_geom_col} -->
<!-- articles_afinn |> 
  filter(section %in% c("News", "Sport")) |> 
  group_by(section, value) |> 
  summarise(sentiment = sum(value)) |> 
  ungroup() |>
  ggplot(mapping = aes(x = value, y = sentiment, fill = section)) +
  geom_col(position = "dodge") -->

<!-- ``` -->



<!-- ```{r articles_afinn_ggplot_word_president} -->
<!-- articles_afinn |> 
  count(section, word, value, sort = TRUE) |> 
  ungroup() |> 
  group_by(section, value) |> 
  slice_max(n, n = 3) |> 
  ungroup() |> 
  mutate(word = reorder(word, n)) |> 
  ggplot(mapping = aes(n, word, fill = section)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~value, scales = "free_y") +
  labs(x = "Contribution to sentiment", 
       y = NULL) -->
<!-- ``` -->

# n-grams and correlations

So far we have been looking at words as individual units, and not considered how they are related to the other words around it. It is possible to make analysis where you look at the relationsships between word in our text.

Instead of using the `unnest_tokens` function to by word, as we have done so far, we will tokenize our text in to sequences of words called n-grams. This gives us the possibility to see how often a word is followed by another word, and hereby gives us the chance to look at the relationsship between words.

We no longer wants to use `articles_filteres` since this is tokenized by word. We have a `dataframe` that contains the articles text in one cell, so we will go back to our orginal object `articles` that contain all the articles.

So lets tokenize to 2-words.

 ```{r}
articles_bigrams <- articles |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |> 
  filter(!is.na(bigram))
 ```

```{r}
articles_bigrams
```

It looks much like the result of a tokenisation by word, but the added column is now called bigram and contains two words

We can now count how many times word pair occurs.

```{r}
articles_bigrams |> 
  count(bigram, sort = TRUE)
```

Here we can see that the birams that tops the list are pairs of quite common words, much of these words are the once we earlier called stopwords. It would be nice to remove the pair where one of the words are a stopword.

In order to be able to remove these pairs so we have to put each word in it own column. We can do this by using the function `separate`.

First we will separate the pair into two columns by separating the pair around the space between them.

```{r}
bigrams_separated <- articles_bigrams |> 
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_separated
```

After that we will remove the rows that contain a stopword in either of the two new columns.

```{r}
bigrams_filtered <- bigrams_separated |> 
  filter(!word1 %in% stop_words$word) |> 
  filter(!word2 %in% stop_words$word)

bigrams_filtered
```

Now we can make a new count of word pair.
```{r}
bigram_counts <- bigrams_filtered |> 
  count(word1, word2, sort = TRUE)

bigram_counts
```

Now we get some more meaning full word pairs.

If we want to combine the colums again to have the words pairs (without stopwords) in one column, it can easily be done by using the function `unite`

```{r}
bigrams_united <- bigrams_filtered |> 
  unite(bigram, word1, word2, sep = " ")

bigrams_united
```

We could have a look at how different word pairs, are used in different sections.

```{r}
bigrams_united |> 
  count(section, bigram, sort = TRUE) |> 
  pivot_wider(
    names_from = section,
    values_from = n)

```




::::::::::::::::::::::::::::::::::::: keypoints 

- Sentiments is the emotion or tone in a text
- There are different lexicons
- It is possible to add sentiments to words
- It is possible to visualise the sentiments
- It is possible to look at relationsship between words

::::::::::::::::::::::::::::::::::::::::::::::::
